---
title: "Climate Futures"
author: "Amber Runyon"
date: "3/23/2022"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here); library(plyr); # Use here::here when package lubridate is used
library(plotrix); library(zoo); library(ggplot2); library(grid); library(cowplot); library(reshape2); library(raster); library(ncdf4); library(reshape2); library(WriteXLS); library(data.table); library(RColorBrewer); library(ggrepel); library(lubridate); library(dplyr); library(gridExtra); library(tidyr); library(tibble); library(stringr); library(ggpubr); library(lemon)

```

## Climate Futures

When trying to adapt to a changing climate, with all the inherent uncertainties about how the future may play out, resource managers often turn to scenario planning as a tool.  Managers use scenario planning to explore plausible ways the climate may change, allowing them to work with climate change uncertainty rather than being paralyzed by it.  Once identified, scenarios of the future are used to develop proactive measures to prepare for and adapt to scenarios of change.   A key part of scenario planning is generating a list of potential future climates we may experience. This script uses methods described in [Lawrence et al. (2020)](https://link.springer.com/epdf/10.1007/s10584-021-03169-y?sharing_token=MufjWsIQwncLzXdQ2awm9Pe4RwlQNchNByi7wbcMAY48-OL9-_d-75TukkQDtC_QgmuzZqcD52_phTlMsDeKUZ3hlN2p-e60d1ZHtYKyTYClg80yAvWfomsQIPT76_NhyeJSwspZWfdbrpOv4l1egfAD8shVTBGLvZL2_N5-GaQ%3D) and [Runyon et al. (2019)](https://escholarship.org/uc/item/76p7m8rz) to identify divergent, plausible climate futures for selected national parks.

## Steps to run the code
This code uses CMIP5 gloabal climate projections (2005-2099) downscaled by the MACA (Multivariate Adaptive Constructed Analogs) method (Abatzoglou and Brown 2012) with the METDATAv2 (Abatzoglou 2011) observational dataset as training data. (See <https://climate.northwestknowledge.net/MACA/MACAmethod.php> for details). The historical observational data used is GridMET, which is the training data used for the MACA downscaleing (Abatzoglou 2013, <https://climate.northwestknowledge.net/MACA/MACAtrainingdata.php>).

To run this code, you must first enter the 4-letter code for the park unit you wish to plot (e.g., Rocky Mountain National Park is ROMO) and the location where you would like this data saved once it is downloaded. The following chunk of code will then download the historical and projected climate data from the MACA and GridMET datasets. The projected data is daily for 20 GCMs and 2 RCPs, so it may take a second. 


```{r download-data, echo = FALSE, message=FALSE, warning=FALSE}
SiteID = "ROMO" 
DataDir = paste0('C:/Users/achildress/Documents/RCF_Testing/',SiteID)
if(dir.exists(DataDir) == FALSE){
  dir.create(DataDir)
}
# Read CSV files from WB project downloaded from:
  #  https://parkfutures.s3.us-west-2.amazonaws.com/maca-tprh-data/index.html

#################   Functions  ##################     
checkFile <- function(fName, fileURL=F){
  tmpFiles <- list.files(here::here(DataDir))
  if(fName %in% tmpFiles){
    print(paste(fName, " File exists. Not downloaded"))
    if("zip" == tolower(str_sub(fName, str_length(fName)-2, str_length(fName))))unzip(paste(DataDir,fName,sep='/')) 
    return(1)
  }  # 1 file exists
  # no file so go get it
  print(paste("####  retrieving ", fName, " data file. Might take a while.  ######"))
  options(timeout = 300)   # seconds. How long are you willing to wait?
  getName <- paste("https://parkfutures.s3.us-west-2.amazonaws.com/maca-tprh-data/",fName, sep='')
  
  if(fileURL!=F)getName <- paste(fileURL, fName, sep='')
  download.file(getName, destfile=paste(DataDir,fName,sep='/'), mode="wb")
  # unzip if necessary
  if("zip" == tolower(str_sub(fName, str_length(fName)-2, str_length(fName))))unzip(paste(DataDir,fName,sep='/'))
}   # end checkFile


TFtoC <- function(T){(T-32)/1.8}

## Get Data

histInFile <-  paste(SiteID,"_historical.csv", sep='')        # histInFile <- file.choose()    #  browse to path/file
projInFile <- paste(SiteID,"_future.csv", sep='')
checkFile(histInFile)
Gridmet<- read.csv(paste(DataDir, histInFile, sep='/'))
names(Gridmet) <- c("Date","GCM","PrcpIn","TmaxF","TminF","RHmaxPct","RHminPct","TavgF")
Gridmet <- Gridmet %>% mutate(Year = year(Date),
                              RCP = "Hist",
                              DOY = yday(Date))   # for plotting

checkFile(projInFile)
Future_all <- read.csv(paste(DataDir, projInFile, sep="/"))
Future_all$Year <- year(Future_all$Date)
Future_all$RCP <- str_sub(Future_all$GCM, str_length(Future_all$GCM)-1, str_length(Future_all$GCM))

names(Future_all) <- c("Date","GCM","PrcpIn","TmaxF","TminF","RHmaxPct","RHminPct","TavgF","Year", "RCP")

Future_all <- Future_all %>% #dplyr::filter(GCM %in% wbGCMs) %>%
  mutate(DOY = yday(Date))

```

## Select divergent climate futures

Mid-century (2055) change values are calculated for each of the 40 projections (20 GCMs for 2 RCPs; 2040-2070 vs historical [1979-2012]) in the code below and the four projections that represent the most divergent climate futures are selected. Methods for this selection are outlined in Lawrence et al. (2020). 

Note CCRP has produced an R package that conducts all of the following steps. It will be released in early-Summer 2022. 

```{r climate future selection, echo=FALSE}
## Select GCMs
Gridmet$Date = strptime(Gridmet$Date, "%Y-%m-%d")
Future_all$Date = strptime(Future_all$Date, "%Y-%m-%d")

# # Subset Future_all to only be near future (2025-2055) and Baseline_all to only but until 2000
Baseline_all<-Gridmet
Baseline_all<-subset(Baseline_all,Year<2013)

ALL_FUTURE<-Future_all  
Future_all = subset(Future_all, Year >= 2040 & Year <= 2070)
################################# SUMMARIZE CHANGE IN FUTURE TEMP/PRECIP MEANS BY GCM ####################
####Set Average values for all four weather variables, using all baseline years and all climate models
BaseMeanPr = mean(Baseline_all$PrcpIn)
BaseMeanTmx = mean(Baseline_all$TmaxF)
BaseMeanTmn = mean(Baseline_all$TminF)

####Create Future/Baseline means data tables, with averages for all four weather variables, organized by GCM
Future_Means = aggregate(cbind(PrcpIn, TmaxF, TminF, TavgF)
                                    ~ GCM, Future_all, mean,na.rm=FALSE)   # , Future_all$Wind
# names(Future_Means) = c("GCM", "PrcpIn", "TmaxF", "TminF", "TavgF")    # , "Wind"

Baseline_Means = aggregate(cbind(PrcpIn, TmaxF, TminF, TavgF)~GCM, 
                                      Baseline_all, mean)   
# names(Baseline_Means) = c("GCM", "PrcpIn", "TmaxF", "TminF", "TavgF") 

#### add delta columns in order to classify CFs
Future_Means$DeltaPr = Future_Means$PrcpIn - Baseline_Means$PrcpIn
Future_Means$DeltaTmx = Future_Means$TmaxF - Baseline_Means$TmaxF
Future_Means$DeltaTmn = Future_Means$TminF - Baseline_Means$TminF
Future_Means$DeltaTavg = Future_Means$TavgF - Baseline_Means$TavgF

#### Set limits for CF classification
Pr0 = as.numeric(quantile(Future_Means$DeltaPr, 0))
Pr25 = as.numeric(quantile(Future_Means$DeltaPr, 0.25))
PrAvg = as.numeric(mean(Future_Means$DeltaPr))
Pr75 = as.numeric(quantile(Future_Means$DeltaPr, 0.75))
Pr100 = as.numeric(quantile(Future_Means$DeltaPr, 1))
Tavg0 = as.numeric(quantile(Future_Means$DeltaTavg, 0))
Tavg25 = as.numeric(quantile(Future_Means$DeltaTavg, 0.25)) 
Tavg = as.numeric(mean(Future_Means$DeltaTavg))
Tavg75 = as.numeric(quantile(Future_Means$DeltaTavg, 0.75))
Tavg100 = as.numeric(quantile(Future_Means$DeltaTavg, 1))

#### Designate Climate Future
Future_Means$CF1 = as.numeric((Future_Means$DeltaTavg<Tavg & Future_Means$DeltaPr>Pr75) | Future_Means$DeltaTavg<Tavg25 & Future_Means$DeltaPr>PrAvg)
Future_Means$CF2 = as.numeric((Future_Means$DeltaTavg>Tavg & Future_Means$DeltaPr>Pr75) | Future_Means$DeltaTavg>Tavg75 & Future_Means$DeltaPr>PrAvg)
Future_Means$CF3 = as.numeric((Future_Means$DeltaTavg>Tavg25 & Future_Means$DeltaTavg<Tavg75) & (Future_Means$DeltaPr>Pr25 & Future_Means$DeltaPr<Pr75))
Future_Means$CF4 = as.numeric((Future_Means$DeltaTavg<Tavg & Future_Means$DeltaPr<Pr25) | Future_Means$DeltaTavg<Tavg25 & Future_Means$DeltaPr<PrAvg)
Future_Means$CF5 = as.numeric((Future_Means$DeltaTavg>Tavg & Future_Means$DeltaPr<Pr25) | Future_Means$DeltaTavg>Tavg75 & Future_Means$DeltaPr<PrAvg)


#Assign full name of climate future to new variable CF
Future_Means$CF[Future_Means$CF1==1]="Warm Wet"
Future_Means$CF[Future_Means$CF2==1]="Hot Wet"
Future_Means$CF[Future_Means$CF3==1]="Central"
Future_Means$CF[Future_Means$CF4==1]="Warm Dry"
Future_Means$CF[Future_Means$CF5==1]="Hot Dry"

#     Remove extraneous Climate Future columns
Future_Means$CF1 = NULL
Future_Means$CF2 = NULL
Future_Means$CF3 = NULL
Future_Means$CF4 = NULL
Future_Means$CF5 = NULL

#### Select Corner GCMs
lx = min(Future_Means$DeltaTavg)
ux = max(Future_Means$DeltaTavg)
ly = min(Future_Means$DeltaPr)
uy = max(Future_Means$DeltaPr)

  #convert to points
ww = c(lx,uy)
wd = c(lx,ly)
hw = c(ux,uy)
hd = c(ux,ly)

pts <- Future_Means

  #calc Euclidian dist of each point from corners
pts$WW.distance <- sqrt((pts$DeltaTavg - ww[1])^2 + (pts$DeltaPr - ww[2])^2)
pts$WD.distance <- sqrt((pts$DeltaTavg - wd[1])^2 + (pts$DeltaPr - wd[2])^2)
pts$HW.distance <- sqrt((pts$DeltaTavg - hw[1])^2 + (pts$DeltaPr - hw[2])^2)
pts$HD.distance <- sqrt((pts$DeltaTavg - hd[1])^2 + (pts$DeltaPr - hd[2])^2)

pts %>% filter(CF == "Warm Wet") %>% slice(which.min(WW.distance)) %>% .$GCM -> ww
pts %>% filter(CF == "Warm Dry") %>% slice(which.min(WD.distance)) %>% .$GCM -> wd
pts %>% filter(CF == "Hot Wet") %>% slice(which.min(HW.distance)) %>% .$GCM -> hw
pts %>% filter(CF == "Hot Dry") %>% slice(which.min(HD.distance)) %>% .$GCM -> hd

Future_Means %>% mutate(corners = ifelse(GCM == ww,"Warm Wet",
                                         ifelse(GCM == wd, "Warm Dry",
                                                ifelse(GCM == hw, "Hot Wet",
                                                       ifelse( GCM == hd, "Hot Dry",NA))))) -> Future_Means
```

The four individual models selected as climate futures for "r SiteID" are:
```{r four-cf-table, echo=FALSE}
Climate.futures <- Future_Means %>% drop_na() %>% select(c(GCM,CF))
knitr::kable(Climate.futures, format="markdown")
```

These models are plotted relative to the other models. 
```{r scatterplot,echo=FALSE, warning=FALSE}

dualscatter = ggplot(Future_Means, aes(DeltaTavg, DeltaPr*365, xmin=Tavg25, xmax=Tavg75, ymin=Pr25*365, ymax=Pr75*365))

dualscatter  + geom_text_repel(aes(label=GCM)) +
  geom_point(colour="black",size=4) +
 geom_point(aes(x=DeltaTavg[which(corners=="Warm Wet")], y=365*DeltaPr[which(corners=="Warm Wet")]), shape=21, size=10, stroke=3, colour="cornflowerblue") +
   geom_point(aes(x=DeltaTavg[which(corners=="Hot Dry")], y=365*DeltaPr[which(corners=="Hot Dry")]), shape=21, size=10, stroke=3, colour="red") +
     geom_point(aes(x=DeltaTavg[which(corners=="Warm Dry")], y=365*DeltaPr[which(corners=="Warm Dry")]), shape=21, size=10, stroke=3, colour="gold1") +
     geom_point(aes(x=DeltaTavg[which(corners=="Hot Wet")], y=365*DeltaPr[which(corners=="Hot Wet")]), shape=21, size=10, stroke=3, colour="green4") +

  theme(axis.text=element_text(size=18),
        axis.title.x=element_text(size=18,vjust=-0.2),
        axis.title.y=element_text(size=18,vjust=0.2),
        plot.title=element_text(size=18,face="bold",vjust=2,hjust=0.5),
        legend.text=element_text(size=18), legend.title=element_text(size=16)) + 
  ###
  labs(title =paste(SiteID," Changes in climate means \n (2020-2099 vs 1979-2012) by GCM run",sep=""), 
       x = "Changes in annual average temperature (\u00B0F)", # Change
       y = "Changes in annual average precipitation (in)") + #change
  scale_color_manual(name="Scenarios", values=c("black")) +
  # scale_fill_manual(name="Scenarios",values = c("black")) + 
  theme(legend.position="none") 
  

```


Then annual temperature can be summarized for the cliamte futures. Here I have selected the "warm wet" and "hot dry" climate futures as these typically represent the "best" and "worst" case scenarios for park resources. 

```{r plot annual temperature, echo=FALSE}
degF <- "(\u00B0F)"

## Plotting functions
#ggplot theme to control formatting parameters for plots with month on the x-axis
PlotTheme = theme(axis.text=element_text(size=20),    #Text size for axis tick mark labels
                  axis.title.x=element_blank(),               #Text size and alignment for x-axis label
                  axis.title.y=element_text(size=24, vjust=0.5,  margin=margin(t=20, r=20, b=20, l=20)),              #Text size and alignment for y-axis label
                  plot.title=element_text(size=26,face="bold",hjust=0.5, margin=margin(t=20, r=20, b=20, l=20)),      #Text size and alignment for plot title
                  legend.title=element_text(size=24),                                                                    #Text size of legend category labels
                  legend.text=element_text(size=22),                                                                   #Text size of legend title
                  legend.position = "bottom")  

LT_plot <-function(data,yvar, rollvar,cols,yaxis,title){
  ggplot(data, aes(x=Year, y={{yvar}}, col=CF, fill=CF)) + 
    # geom_ribbon(aes(x=as.numeric(as.character(year)), ymin=Tavg.min, ymax=Tavg.max, fill=CF), alpha=0.5) +
    geom_line(size=2) + geom_point(col="black", size=2, shape=16) +
    geom_point() +
    geom_line(aes(x=Year, y={{rollvar}}),size=1.25,colour="black", na.rm=TRUE) +
    geom_line(aes(x=Year, y={{rollvar}},colour = CF), size=.75 ,na.rm=TRUE) +
    scale_x_continuous(breaks=c(1980, 2000, 2020, 2040, 2060, 2080, 2100)) +
    labs(x="Year", y=yaxis,title=title) +
    scale_color_manual(name="Climate Future",values=cols) +
    scale_fill_manual(name="Climate Future",values=cols) + PlotTheme +
    theme( axis.line = element_line(colour = "black"), #Add axis lines
           panel.background = element_blank(), #Background white
           panel.grid.major = element_line("light grey",0.3)) #add grid back)
}

## Format and summarize data
Future_summary <- ALL_FUTURE %>% left_join(Climate.futures,by="GCM")
Baseline_summary <- Gridmet; Baseline_summary$CF = "Historical"
all_summary <- rbind(Baseline_summary, Future_summary)
all_summary <- subset(all_summary, CF %in% c("Warm Wet", "Hot Dry", "Historical"))
all_summary$CF <- factor(all_summary$CF, levels=c("Historical", "Warm Wet","Hot Dry"))

yrAvgs <- aggregate(cbind(TavgF, PrcpIn)~Year+CF,all_summary,mean)
yrAvgs$PrcpIn <- yrAvgs$PrcpIn * 365
yrAvgs$TavgRoll10 <- rollmean(yrAvgs$TavgF, 10, fill = NA, align = "right")
yrAvgs$PrcpRoll10 <- rollmean(yrAvgs$Prcp, 10, fill = NA, align = "right")

## Plots
  # tmean
LT_plot(yrAvgs,TavgF,rollvar=TavgRoll10,cols=c("darkgray","cornflowerblue","red"),yaxis="Mean annual temperature (\u00B0F)",title="") 

  # Precip
LT_plot(yrAvgs,PrcpIn,rollvar=PrcpRoll10,cols=c("darkgray","cornflowerblue","red"),yaxis="Mean annual precipitation (inches/Yr)",title="")

```
